---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# arbitree

<!-- badges: start -->
<!-- badges: end -->

## Installation

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("manzt/arbitree")
```
## Usage

Arbitree is intended to plugin to the normal Monocle3 workflow. Users are encouraged to read the [docs](https://cole-trapnell-lab.github.io/monocle3/docs/trajectories/) prior to performing trajectory analysis. We've wrapped the `learn_graph` function from Monocle3 with `arbitree_learn_graph`, which exposes assumed, hard-coded, and hueristic parameters for determinig inital "landmark cells" for DDRTree. Calling `arbitree_learn_graph` without any arguments gives the same result as `learn_graph`.

```{r example, eval = FALSE}
library(monocle3)
library(arbitree) 

cds <- new_cell_data_set(expression_matrix, cell_metadata, gene_metadata)
##
# perform  normalization, gene selection, dimensionality reduction, and clustering...
##
cds <- learn_graph(cds) # monocle function
cds <- arbitree_learn_graph(cds) # arbitree function (same output)

plot_cells(cds) # plot trajectory
```

As described the [paper](https://www.nature.com/articles/s41586-019-0969-x), "[Monocle3] first selects a set of ‘landmark’ cells using by first running the `kmeans()` clustering algorithm in R with k equal to the value of the `ncenter` argument, which can be passed to `learn_graph()` by the user. The landmark cells are then selected by first mapping each cell to its nearest kmeans point, and then selecting the cell for each kmeans point with the highest local density. By default, Monocle 3 uses a data-dependent policy for adjusting ncentre automatically."

In `arbitree_learn_graph`, we have added `ncenter`, `k_nn`, and `init_medioids` as parameters. 

- `ncenter`: number of kmeans clusters ultimately landmark cells.
- `k_nn`: number of nearest neighbors used to determine the landmark cell with highest local density. In Monocle 3, this parameter is hard-coded to 25 and does not scale with the size of the partition.
- `input_medioids`: an (`ndim x centroid`) matrix used to seed the DDRTree algorithm. Default is `NULL`. If provided, the landmark cells used to initialize DDRTree are determined by the cell which is nearest in low dimensional euclidean distance to each provided centroid, skipping the kmeans and knn. 

We created an widget for drawing arbitrary "iniital states" over the reduced dimensions using the `draw_tree` function. Usage is as follows: 

```{r example2, eval = FALSE}
xy_coords <- draw_tree(cds) # opens shiny app and user draws points on UMAP projection
cds <- arbitree_learn_graph(cds, input_medioids = t(xy_coords))
plot_cells(cds)
```

